Binning Strategies and Related Loss for Graphics of Large Data

### 2013-03-06 Paper.Rnw
Trying out GIT
next commit
Fixing pictures, etc. May depend on newest dbData package from github.

@@

\begin{document}
\SweaveOpts{concordance=TRUE, warning=FALSE,message=FALSE}
\maketitle

@@

\section{Introduction}
<<setup,echo=F,include=F,eval=T>>=
library(ggplot2)
library(gridExtra)
library(dbData)
#library(multicore)
library(devtools)
library(xtable)
library(RColorBrewer)

@@

  res <- data.frame(sapply(1:(k-1), function(var) {
			return(rep(x[,var], x[,k]))
		}
	))
	names(res) <- names(x)[1:(k-1)]
	return(res)
}
d1.ext <- xtd.data(d1)
@
<<scatterplots, dependson='setup', echo=F, results='hide'>>=
qplot(data=d1.ext, x=G, y=SO, geom="point", main="Games and Strike Outs in ML Baseball", alpha=I(.05), size=I(3)) + theme_bw()
  ggsave("./figure/OverplottingAlpha.pdf", width=5, height=5)
qplot(data=d1, x=G, y=SO, geom="point", main="Games and Strike Outs in ML Baseball", size=I(3)) + theme_bw()
  ggsave("./figure/Overplotting.pdf", width=5, height=5)
qplot(data=d1, x=G, y=SO, geom="point", main="Games and Strike Outs in ML Baseball", shape="o", size=I(6)) + scale_shape_identity() + theme_bw()
  ggsave("./figure/OverplottingCircles.pdf", width=5, height=5)
qplot(data=d1.ext, x=G, y=SO, geom="hex", fill=..count.., group=1) +  
  scale_fill_gradient("Frequency", low="#56B1F7", high="#132B43", guide="legend", trans="log", breaks=c(1, 10, 100, 1000)) + theme_bw() + theme(legend.position="bottom")
  ggsave("./figure/HexBinning.pdf", width=5, height=5)
d1.hex.obj <- hexbin(d1.ext$G, d1.ext$SO)
d1.hex <- data.frame(hcell2xy(d1.hex.obj))
names(d1.hex) <- c("G", "SO")
d1.hex$Freq <- d1.hex.obj@count
qplot(data=d1.hex, x=G, y=SO, geom="point", size=Freq) + 
  scale_area("Frequency", range=c(1, 4), trans="log", breaks=c(1, 10, 100, 1000, 10000))+ theme_bw() +
  theme(legend.position="bottom")
  ggsave("./figure/BinScatter.pdf", width=5, height=5)

# Sunflower plot
# d1.hex$id <- seq(1:nrow(d1.hex))
# d1.hex2 <-  ddply(d1.hex, .(G, SO, id), function(i) return(data.frame(r = cos((floor(i$Freq/10))*seq(0, 2*pi, .05)), theta=seq(0, 2*pi, 0.05), Freq=i$Freq)))
# ggplot(d1.hex2) + geom_subplot(aes(x=G, y=SO, group=id,  subplot=geom_star(aes(r=r, angle=theta, fill=log(Freq), colour=log(Freq)), r.zero=TRUE)))

@@

Figure~\ref{fig:BinningTutorial} provides an illustration of the binning process.
<<binningDemo,echo=F, results='none', cache=T>>=
      library(mvtnorm)
      set.seed(33)
      d <- rmvnorm(200, c(5, 25), matrix(c(100, 50, 50, 100), nrow=2))
      d <- data.frame(d)
      names(d) <- c("x", "y")
      d <- subset(d, (x>-20 & x<40 & y>0 & y<50),drop=TRUE)
      d.rand <- d
      dnew <- binStd(cbind(d-5, Freq=1), c(10, 10))
      dnew$x <- dnew$x+5
      dnew$y <- dnew$y+5
      d$bin <- dnew$x%in%c(5,15) & dnew$y==25
      d$a <- .33 + .66*d$bin
      d.orig <- d

@@

\subsection{Random Binning Algorithm}
<<RandomBinDemo,echo=F,include=FALSE,results='hide',  cache=TRUE>>=
nl <- ncol(d.rand)
dnew <- d.rand
binning <- c(10, 10)
for(x in 1:(nl)){
  y <- d.rand[,x] - binning[x]/2
dnew[,x] <- binning[x]*(floor(y/binning[x]) + 
  sapply((y%%binning[x])/binning[x], function(p) rbinom(1, 1, p))) + binning[x]/2
}
temp <- data.frame(   
          x=c(-20, -10,   0,  10,  20,  30,  40, -20, -20, -20, -20, -20, -20), 
       xend=c(-20, -10,   0,  10,  20,  30,  40,  40,  40,  40,  40,  40,  40), 
          y=c(  0,   0,   0,   0,   0,   0,   0,   0,  10,  20,  30,  40,  50), 
       yend=c( 50,  50,  50,  50,  50,  50,  50,   0,  10,  20,  30,  40,  50), 
      group=c(  1,   2,   3,   4,   5,   6,   7,   8,   9,  10,  11,  12,  13))

@@

loss0.1 <- data.frame(with(loss0.1, c(Loss, rows=nrow(ddply(NewData, .(x,y), summarise, Freq=sum(fsum))), cols=ncol(ddply(NewData, .(x,y), summarise, Freq=sum(fsum))))))
loss0.25 <- data.frame(with(loss0.25, c(Loss, rows=nrow(ddply(NewData, .(x,y), summarise, Freq=sum(fsum))), cols=ncol(ddply(NewData, .(x,y), summarise, Freq=sum(fsum))))))
loss0.5 <- data.frame(with(loss0.5, c(Loss, rows=nrow(ddply(NewData, .(x,y), summarise, Freq=sum(fsum))), cols=ncol(ddply(NewData, .(x,y), summarise, Freq=sum(fsum))))))
loss1 <- data.frame(with(loss1, c(Loss, rows=nrow(ddply(NewData, .(x,y), summarise, Freq=sum(fsum))), cols=ncol(ddply(NewData, .(x,y), summarise, Freq=sum(fsum))))))

@@

Fortunately we can show that the total loss perceived is partitioned by these two components of loss, and it holds (see appendix \ref{proof:partition} for a proof):
\begin{eqnarray}\label{eqn:partition}
L_s(X) =  1/S^2_{\emptyset} \cdot \left(\sum_i L^N_i + \sum_i L^V_i\right)
\end{eqnarray}
<<smoothing, echo=FALSE, include=FALSE, out.width="0.5\\textwidth", fig.width=5.4, fig.height=4.05>>=
set.seed(53022)
data <- data.frame(x=rnorm(200, 0, 2), y=rnorm(200, 0, 2), Freq=1)
data$xbin <- round(data$x)
data$ybin <- round(data$y)
data.means <- ddply(data, .(xbin,ybin), summarise, xmean=mean(x), ymean=mean(y))
data <- merge(data, data.means)

ggplot()+geom_histogram(aes(x=data$x, y=..count..), binwidth=1/4, fill=I("grey80"), colour=I("grey80"), alpha=I(.5)) + geom_histogram(aes(x=data$x, y=..count../4), binwidth=1, fill=NA, colour=I("grey50"), alpha=I(.5)) + theme_bw() + xlab("x") + ylab("Frequency") + ggtitle("Frequency Loss")
ggsave("./figure/FreqLoss.pdf", width=5.4*.8, height=4.05*.8)
@

@@

For uniformly distributed data (right column) the choice of color scheme does not have a huge impact, but under both the normal and exponential distribution, log frequency scaling results in categories which account for more uniform proportions of the data (as can be seen by the color strips in the second and third row of Figure~\ref{fig:logcolor}). This is desirable because it provides more discernible frequency information. 
\begin{figure}[hbtp]
\centering
<<FrequencyCategories,echo=FALSE, out.width='\\textwidth', fig.width=8.5, fig.height=11>>=
options(stringsAsFactors=FALSE)
library(plyr)
library(ggplot2)
library(RColorBrewer)
set.seed(5032)
data <- rbind(data.frame(x=runif(1000, 0, 20), y=runif(1000, 0, 20), 
                         Freq=1, Distribution="Uniform"), 
              data.frame(x=rnorm(1000, 10, 3), y=rnorm(1000, 10, 3), 
                         Freq=1, Distribution="Normal"),
              data.frame(x=rexp(1000, .3), y=rexp(1000, .3), 
                         Freq=1, Distribution="Exponential"))
data <- data[which(abs(data$x-10)<10 & abs(data$y-10)<10),]
# qplot(data=data, x=x, y=y, geom="point")+facet_wrap(~Distribution)

@@

X <- data.frame(fill=with(linearbins, rep(fill, Freq)), Distribution = with(linearbins, rep(Distribution, Freq)))
X <- ddply(X, .(Distribution), transform, id=1:length(Distribution))
p2 <- qplot(id, binwidth=1, fill=fill, data=X, alpha=I(1)) + 
  scale_fill_manual(values=colors, name="Frequency", guide="none") + 
  facet_grid(.~Distribution, scales="free") +
  xlab("Count") + theme_bw() + ylab("") +
  ggtitle("(b) Color Frequency: Linear Binning") + 
  theme(axis.text.y=element_text(colour="white"), axis.ticks=element_blank()) + 
  theme(plot.title = element_text(hjust=0, lineheight=.8, size=12), plot.margin = unit(c(0,0,0,-0.35), "cm"))
# linear binning
pdf(file="figure/p2.pdf", width=8, height=3)
p2
dev.off()

@@

multiplot(p1, p2, p4, p3, cols=1, heights=c(2.2,1,1,2.2))
#ggsave("./figure/MultipleDensityFigure.png", width=8.5, height=11, dpi=1200)
@
%\includegraphics[width=\textwidth]{figure/MultipleDensityFigure}
\caption{\label{fig:logcolor} Linear and Log Frequency Binning  to Simulated Data from Uniform, Normal, and Exponential distributions}
\end{figure}

@@

\section{Discussion and Examples}
\subsection{Comparison of Binning Algorithms}
<<LossTableGraphs,echo=F,results="hide",cache=TRUE,fig.keep='none', eval=FALSE>>=
LossTable <- NULL
# bins <- rbind(c(1,1), c(2,2), c(2,5), c(5, 2), c(4,4), c(3, 5), c(5, 3), c(5, 5), c(5, 25), c(25, 5), c(2, 50), c(50, 2), c(5, 50))
bins <- rbind(c(2,2), c(5,25))
computeLoss <- function(binning, data){
  lRandom <- lossCalc(data, binning=binning, type="random", newData = TRUE)
  lMean <- lossCalc(data, binning=binning, type="standard", newData=TRUE)
  datplot <- rbind(cbind(lRandom$NewData, Algorithm="Random"), cbind(lMean$NewData, Algorithm="Standard"))
    qplot(G, SO, fill = Freq, geom="tile", data = datplot, 
        main=paste("Baseball: Binning by (", binning[1], ", ", binning[2],")", sep="")) + facet_grid(.~Algorithm) +
        scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend", trans="log", breaks=c(1, 5, 50, 500, 5000)) + theme_bw() + theme(legend.position="bottom")
    ggsave(paste("./figure/RandomStd", binning[1], binning[2], ".pdf", sep=""), width=8, height=5, units="in", dpi=1200)
#     qplot(G, SO, fill = log10(Freq), geom="tile", data = lMean$NewData, 
#       main=paste("Standard Binning (", binning[1], ", ", binning[2],")", sep="")) + 
#       scale_fill_gradient(low="#56B1F7", high="#132B43", guide="none", trans="log") + theme_bw()
#     ggsave(paste("./figure/Standard", binning[1], binning[2], ".pdf", sep=""), width=4.5, height=4.5, units="in")
  return(rbind(
    cbind(binX = binning[1], binY = binning[2], Function = "Random", lRandom$LossAll), 
    cbind(binX = binning[1], binY = binning[2], Function = "Standard", lRandom$LossAll)))
}

LossTable <- do.call("rbind", lapply(1:nrow(bins), function(i) computeLoss(bins[i,], d1)))
LossTable <- as.data.frame(LossTable)
names(LossTable) <- c("binX", "binY", "Function", "G", "SO", "Freq")
tab <- xtable(LossTable[order(LossTable[,3], LossTable[,1]),], digits=c(0, 0, 0, 0, 8, 8, 8))
@

@@

% \begin{subsection}{Numerical Comparison of Binning Algorithms}
<<PrintTable,echo=F,include=FALSE>>=
  print(tab, include.rownames=FALSE)
@
% \end{subsection}

>>>>

%Karsten was here
%Heike, too

@@

\begin{document}
% \SweaveOpts{concordance=TRUE, warning=FALSE,message=FALSE}
\maketitle

@@

\section{Introduction}
<<setup,echo=F,include=F,eval=T>>=
library(ggplot2)
library(gridExtra)
library(dbData)
library(devtools)
library(xtable)
library(RColorBrewer)
opts_chunk$set(concordance=TRUE, tidy=TRUE, cache=TRUE, message=FALSE, warning=FALSE)

@@

  res <- data.frame(sapply(1:(k-1), function(var) {
			return(rep(x[,var], x[,k]))
		}
	))
	names(res) <- names(x)[1:(k-1)]
	return(res)
}
d1.ext <- xtd.data(d1)
@
<<scatterplots, dependson='setup', echo=F, include=FALSE>>=
library(hexbin)
qplot(data=d1.ext, x=G, y=SO, geom="point", main="Games and Strike Outs in ML Baseball", alpha=I(.05), size=I(3)) + theme_bw()
  ggsave("./figure/OverplottingAlpha.pdf", width=5, height=5)
qplot(data=d1, x=G, y=SO, geom="point", main="Games and Strike Outs in ML Baseball", size=I(3)) + theme_bw()
  ggsave("./figure/Overplotting.pdf", width=5, height=5)
qplot(data=d1, x=G, y=SO, geom="point", main="Games and Strike Outs in ML Baseball", shape="o", size=I(6)) + scale_shape_identity() + theme_bw()
  ggsave("./figure/OverplottingCircles.pdf", width=5, height=5)
qplot(data=d1.ext, x=G, y=SO, geom="hex", fill=..count.., group=1) +  
  scale_fill_gradient("Frequency", low="#56B1F7", high="#132B43", guide="legend", trans="log", breaks=c(1, 10, 100, 1000)) + theme_bw() + theme(legend.position="bottom")
  ggsave("./figure/HexBinning.pdf", width=5, height=5)
d1.hex.obj <- hexbin(d1.ext$G, d1.ext$SO)
d1.hex <- data.frame(hcell2xy(d1.hex.obj))
names(d1.hex) <- c("G", "SO")
d1.hex$Freq <- d1.hex.obj@count
qplot(data=d1.hex, x=G, y=SO, geom="point", size=Freq) + 
  scale_size_area("Frequency", max_size=4, trans="log", breaks=c(1, 10, 100, 1000, 10000))+ theme_bw() +
  theme(legend.position="bottom")
  ggsave("./figure/BinScatter.pdf", width=5, height=5)

# Sunflower plot
# d1.hex$id <- seq(1:nrow(d1.hex))
# d1.hex2 <-  ddply(d1.hex, .(G, SO, id), function(i) return(data.frame(r = cos((floor(i$Freq/10))*seq(0, 2*pi, .05)), theta=seq(0, 2*pi, 0.05), Freq=i$Freq)))
# ggplot(d1.hex2) + geom_subplot(aes(x=G, y=SO, group=id,  subplot=geom_star(aes(r=r, angle=theta, fill=log(Freq), colour=log(Freq)), r.zero=TRUE)))
@

@@

Figure~\ref{fig:BinningTutorial} provides an illustration of the binning process.
<<binningDemo, echo=FALSE, include=FALSE>>=
      library(mvtnorm)
      set.seed(33)
      d <- rmvnorm(200, c(5, 25), matrix(c(100, 50, 50, 100), nrow=2))
      d <- data.frame(d)
      names(d) <- c("x", "y")
      d <- subset(d, (x>-20 & x<40 & y>0 & y<50),drop=TRUE)
      d.rand <- d
      dnew <- binStd(cbind(d-5, Freq=1), c(10, 10))
      dnew$x <- dnew$x+5
      dnew$y <- dnew$y+5
      d$bin <- dnew$x%in%c(5,15) & dnew$y==25
      d$a <- .33 + .66*d$bin
      d.orig <- d

@@

\subsection{Random Binning Algorithm}
<<RandomBinDemo,echo=F,include=FALSE,cache=TRUE>>=
nl <- ncol(d.rand)
dnew <- d.rand
binning <- c(10, 10)
for(x in 1:(nl)){
  y <- d.rand[,x] - binning[x]/2
dnew[,x] <- binning[x]*(floor(y/binning[x]) + 
  sapply((y%%binning[x])/binning[x], function(p) rbinom(1, 1, p))) + binning[x]/2
}
temp <- data.frame(   
          x=c(-20, -10,   0,  10,  20,  30,  40, -20, -20, -20, -20, -20, -20), 
       xend=c(-20, -10,   0,  10,  20,  30,  40,  40,  40,  40,  40,  40,  40), 
          y=c(  0,   0,   0,   0,   0,   0,   0,   0,  10,  20,  30,  40,  50), 
       yend=c( 50,  50,  50,  50,  50,  50,  50,   0,  10,  20,  30,  40,  50), 
      group=c(  1,   2,   3,   4,   5,   6,   7,   8,   9,  10,  11,  12,  13))

@@

loss0.1 <- data.frame(with(loss0.1, c(Loss, rows=nrow(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))), cols=ncol(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))))))
loss0.25 <- data.frame(with(loss0.25, c(Loss, rows=nrow(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))), cols=ncol(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))))))
loss0.5 <- data.frame(with(loss0.5, c(Loss, rows=nrow(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))), cols=ncol(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))))))
loss1 <- data.frame(with(loss1, c(Loss, rows=nrow(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))), cols=ncol(ddply(NewData, .(x,y), summarise, Freq=sum(Freq))))))

@@

Fortunately we can show that the total loss perceived is partitioned by these two components of loss, and it holds (see appendix \ref{proof:partition} for a proof):
\begin{eqnarray}\label{eqn:partition}
L_s(X) =  1/S^2_{\emptyset} \cdot \left(\sum_i L^N_i + \sum_i L^V_i\right)
\end{eqnarray}

<<smoothing, echo=FALSE, include=FALSE, eval=FALSE>>=
library(plyr)
library(ggplot2)
library(reshape)
set.seed(53022)
data <- data.frame(x=rnorm(200, 0, 2), y=rnorm(200, 0, 2), Freq=1)
data$xbin <- round(data$x)
data$ybin <- round(data$y)
data.means <- ddply(data, .(xbin,ybin), summarise, xmean=mean(x), ymean=mean(y))
data <- merge(data, data.means)

ggplot()+
  geom_histogram(aes(x=data$x, y=..count..), binwidth=1/4, fill=I("grey80"), colour=I("grey80"), alpha=I(.5)) + 
  geom_histogram(aes(x=data$x, y=..count../4), binwidth=1, fill=NA, colour=I("grey50"), alpha=I(.5)) + 
  theme_bw() + xlab("x") + ylab("Frequency") + ggtitle("Frequency Loss")
ggsave("./figure/FreqLoss.pdf", width=5.4*.8, height=4.05*.8)
@

@@

For uniformly distributed data (right column) the choice of color scheme does not have a huge impact, but under both the normal and exponential distribution, log frequency scaling results in categories which account for more uniform proportions of the data (as can be seen by the color strips in the second and third row of Figure~\ref{fig:logcolor}). This is desirable because it provides more discernible frequency information. 
\begin{figure}[hbtp]
\centering
<<FrequencyCategories, echo=FALSE, out.width='\\linewidth', fig.width=8.5, fig.height=11>>=
options(stringsAsFactors=FALSE)
library(plyr)
library(ggplot2)
library(RColorBrewer)
set.seed(5032)
data <- rbind(data.frame(x=runif(1000, 0, 20), y=runif(1000, 0, 20), 
                         Freq=1, Distribution="Uniform"), 
              data.frame(x=rnorm(1000, 10, 3), y=rnorm(1000, 10, 3), 
                         Freq=1, Distribution="Normal"),
              data.frame(x=rexp(1000, .3), y=rexp(1000, .3), 
                         Freq=1, Distribution="Exponential"))
data <- data[which(abs(data$x-10)<10 & abs(data$y-10)<10),]
# qplot(data=data, x=x, y=y, geom="point")+facet_wrap(~Distribution)

@@

X <- data.frame(fill=with(linearbins, rep(fill, Freq)), Distribution = with(linearbins, rep(Distribution, Freq)))
X <- ddply(X, .(Distribution), transform, id=1:length(Distribution))
p2 <- qplot(id, binwidth=1, fill=fill, data=X, alpha=I(1)) + 
  scale_fill_manual(values=colors, name="Frequency", guide="none") + 
  facet_grid(.~Distribution, scales="free") +
  xlab("Count") + theme_bw() + ylab("") +
  ggtitle("(b) Color Frequency: Linear Binning") + 
  theme(axis.text.y=element_text(colour="white"), axis.ticks=element_blank()) + 
  theme(plot.title = element_text(hjust=0, lineheight=.8, size=12), plot.margin = unit(c(0,0,0,-0.35), "cm"))
# linear binning
# pdf(file="figure/p2.pdf", width=8, height=3)
# p2
# dev.off()

@@

multiplot(p1, p2, p4, p3, cols=1, heights=c(2.2,1,1,2.2))
# ggsave("./figure/MultipleDensityFigure.png", width=8.5, height=11, dpi=1200)
@
%\includegraphics[width=\textwidth]{figure/MultipleDensityFigure}
\caption{\label{fig:logcolor} Linear and Log Frequency Binning  to Simulated Data from Uniform, Normal, and Exponential distributions}
\end{figure}

@@

\section{Discussion and Examples}
\subsection{Comparison of Binning Algorithms}
<<LossTableGraphs,echo=F,results="hide",cache=TRUE,fig.keep='none', eval=FALSE>>=
# connect <- dbConnect(dbDriver("MySQL"), user="2009Expo", 
#                    password="R R0cks", port=3306, dbname="baseball", 
#                    host="headnode.stat.iastate.edu")
# pitch <- new("dataDB", co=connect, table="Pitching")
# 
# d1 <- dbData(pitch, vars=c( "G", "SO"))

LossTable <- NULL
bins <- rbind(c(2,2), c(5,25))
computeLoss <- function(binning, data){
  lRandom <- lossCalc(data, binning=binning, type="random", newData = TRUE)
  lMean <- lossCalc(data, binning=binning, type="standard", newData=TRUE)
  datplot <- rbind(cbind(lRandom$NewData, Algorithm="Random"), cbind(lMean$NewData, Algorithm="Standard"))
  qplot(G, SO, fill = Freq, geom="tile", data = datplot, 
      main=paste("Baseball: Binning by (", binning[1], ", ", binning[2],")", sep="")) + 
    facet_grid(.~Algorithm) +
    scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend", trans="log", 
                        breaks=c(1, 5, 50, 500, 5000)) + 
    theme_bw() + theme(legend.position="bottom")
  ggsave(paste("./figure/RandomStd", binning[1], binning[2], ".pdf", sep=""), 
         width=8, height=5, units="in", dpi=1200)

  return(rbind(
    data.frame(binX = binning[1], binY = binning[2], Function = "Random", loss=lRandom$Loss[,5:7]), 
    data.frame(binX = binning[1], binY = binning[2], Function = "Standard", loss=lMean$Loss[,5:7])))
}

LossTable <- do.call("rbind", lapply(1:nrow(bins), function(i) computeLoss(bins[i,], d1)))
LossTable <- as.data.frame(LossTable)
names(LossTable2) <- c("binX", "binY", "Function", "G", "SO", "Freq")
tab <- xtable(LossTable[order(LossTable[,3], LossTable[,1]),], digits=c(0, 0, 0, 0, 8, 8, 8))
@

@@

% \begin{subsection}{Numerical Comparison of Binning Algorithms}
<<PrintTable,echo=F,include=FALSE>>=
#   print(tab, include.rownames=FALSE)
@
% \end{subsection}
